---
title: "Transit-Oriented Development in Boston, MA"
author: "Michael Clifford and Shuai Wang"
date: '2023-09-22'
output: 
  html_document:
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)
library(mapview)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")
```

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("730e334cd384273fdcda9984e41c6444518d82d6", overwrite = TRUE)
```



```{r results='hide'}
#Get census tract data
tracts19 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E", "B25058_001E",
                        "B06012_002E", "B01002_001E"), 
          year=2019, state=25,
          county=025, geometry=TRUE) %>% 
  st_transform('ESRI:102686')

tracts09 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E", "B25058_001E",
                        "B06012_002E", "B01002_001E"), 
          year=2009, state=25,
          county=025, geometry=TRUE) %>% 
  st_transform('ESRI:102686')
```







### Working with ACS Data

```{r}
# Let's "spread" the data into wide form

tracts19 <- 
  tracts19 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         FemaleBachelors = B15001_050, 
         MaleBachelors = B15001_009,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         TotalPoverty = B06012_002,
         MedAge = B01002_001)

tracts09 <- 
  tracts09 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         FemaleBachelors = B15001_050, 
         MaleBachelors = B15001_009,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         TotalPoverty = B06012_002,
         MedAge = B01002_001)


# Let's create new rate variables using mutate

tracts19 <- 
  tracts19 %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2019") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)

tracts09 <- 
  tracts09 %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2009") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)
```


```{r}
# Combine the tracts from the two years
allTracts <- rbind(tracts09,tracts19)
```

### Wrangling Transit Open Data

```{r}

mbta <- st_read("https://hub.arcgis.com/datasets/massgis::mbta-rapid-transit-stops.geojson")

# Remove Silver Line Stops
mbta <- mbta %>% filter(LINE != "SILVER")

mbta <- mbta %>% st_transform(st_crs(tracts19))
```


```{r}
#plot the stops
ggplot() + 
  geom_sf(data=st_union(tracts19)) +
  geom_sf(data=mbta, 
          aes(colour = LINE), 
          show.legend = "point", size= 2) +
  
  labs(title="MBTA Stops", 
       subtitle="Boston, MA", 
       caption="Figure 2.5") +
  mapTheme()
```



```{r}
#create .5 mile buffer from stations
stopBuffer <- st_buffer(mbta, 2640)

stopUnion <- st_union(st_buffer(mbta, 2640))

mbtaBuffers <- 
  rbind(
     stopBuffer %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
     stopUnion %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))
```


```{r}
ggplot() +
  geom_sf(data=mbtaBuffers) +
  geom_sf(data=mbta, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 2.6") +
  mapTheme()
```

Create an sf object with ONLY the unioned buffer from the `septaBuffers`
spatial data frame.

```{r}
buffer_sta <- filter(mbtaBuffers, Legend=="Unioned Buffer")
```

### Spatial Intersection of tracts and buffers

#### Spatial intersection with with `st_centroid()` on polygon centroids



```{r}
selectCentroids <-
  st_centroid(tracts19)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts19, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")
```



```{r}
ggplot() +
  geom_sf(selectCentroids, mapping = aes(fill = TotalPop)) +
  geom_sf(data=mbta, show.legend = "point") +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()
```

### Indicator Maps


```{r}
#join TOD and non-TOD tracts by centroid method
allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2009", MedRent * 1.19, MedRent),
         MedHHInc.inf = ifelse(year == "2009", MedHHInc * 1.19, MedHHInc)) 

```

```{r}
#plot TOD vs non-TOD areas
ggplot() +
  geom_sf(data=allTracts.group, mapping = aes(fill = TOD)) +
  geom_sf(data=mbta)
  mapTheme()
```

2. Visualizations

```{r}
buffer_clipped = st_intersection(buffer, allTracts.group) %>%
  st_union() 


ggplot() +
  geom_sf(data = allTracts.group, aes(fill = q5(MedHHInc.inf))) +
  #geom_sf(data = selectCentroids %>% st_union(), color = "red", linewidth = 1, fill = NA) +
  geom_sf(data = buffer_clipped, color = "red", linewidth = 1, fill = NA) +
  facet_wrap(~year) +
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "MedHHInc.inf"),
                    name = "Med. HH Income\n(Quintile Breaks)") +
  labs(title = "Median HH Income", subtitle = "Boston; 2009-2019 (Adjusted for Inflation)") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))

ggplot() +
  geom_sf(data = allTracts.group, aes(fill = q5(MedRent.inf))) +
  geom_sf(data = buffer_clipped, color = "red", linewidth = 1, fill = NA) +
  facet_wrap(~year) +
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "MedRent.inf"),
                    name = "Med. Rent\n(Quintile Breaks)") +
  labs(title = "Median Rent", subtitle = "Boston; 2009-2019 (Adjusted for Inflation)") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))

ggplot() +
  geom_sf(data = allTracts.group, aes(fill = q5(pctBachelors))) +
  geom_sf(data = buffer_clipped, color = "red", linewidth = 1, fill = NA) +
  facet_wrap(~year) +
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "pctBachelors", 0),
                    name = "Pct. Bachelors Degree\n(Quintile Breaks)") +
  labs(title = "Pct. W/ Bachelors Degree", subtitle = "Boston; 2009-2019") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))

ggplot() +
  geom_sf(data = allTracts.group, aes(fill = q5(MedAge))) +
  geom_sf(data = buffer_clipped, color = "red", linewidth = 1, fill = NA) +
  facet_wrap(~year) +
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "MedAge"),
                    name = "Med. Age\n(Quintile Breaks)") +
  labs(title = "Median Age", subtitle = "Boston; 2009-2019") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))
```

### TOD Indicator Plots (3. Bar Plots)

Let's create small multiple plots We use the "gather" command (look this
one up please) To go from wide to long Why do we do this?? Notice we can
"pipe" a ggplot call right into this operation!

```{r}
#Create summary table first
allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Med_HH_Inc = mean(MedHHInc.inf, na.rm = T),
            Med_Rent = mean(MedRent.inf, na.rm = T),
            Percent_Bach = mean(pctBachelors, na.rm = T),
            Med_Age = mean(MedAge, na.rm = T))
```

```{r}
allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Variable, scales = "free", ncol=5) +
  scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
  labs(title = "Indicator differences across time and space") +
  plotTheme() + theme(legend.position="bottom")
```

### TOD Indicator Tables


(4.Table)

```{r}
allTracts.Summary %>%
  unite(year.TOD, year, TOD, sep = ": ", remove = T) %>%
  gather(Variable, Value, -year.TOD) %>%
  mutate(Value = round(Value, 2)) %>%
  spread(year.TOD, Value) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped") %>%
  footnote(general_title = "\n",
           general = "Table 2.3")
```

5. Graduated Symbol Map



6. Line Graph/MRB

```{r MRB}
mbta_MRB <- multipleRingBuffer(st_union(mbta),
                                maxDistance = 31680,
                                interval =  2640)
```

```{r}
ggplot() +
    geom_sf(data=mbta_MRB) +
    geom_sf(data=mbta, size=1) +
    geom_sf(data=st_union(tracts19), fill=NA, size=1.2) +
    labs(title="Half mile buffers") +
    mapTheme()
```

```{r}

tracts19.rings <- tracts19 %>% 
  select(GEOID, year) %>% 
  st_centroid() %>% 
  st_join(mbta_MRB, join = st_intersects) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(tracts19, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles

tracts09.rings <- tracts09 %>% 
  select(GEOID, year) %>% 
  st_centroid() %>% 
  st_join(mbta_MRB, join = st_intersects) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(tracts09, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles

allTracts.rings <- rbind(tracts09.rings, tracts19.rings)

allTracts.rings.summary <- st_drop_geometry(allTracts.rings) %>%
    group_by(distance, year) %>%
    summarize(Mean_Rent = mean(MedRent, na.rm=T))

ggplot(allTracts.rings.summary,
       aes(distance, Mean_Rent, colour=year)) +
      geom_point(size=3) + 
  geom_line(size=2)

```


