---
title: "Transit-Oriented Development in Boston, MA"
author: "Michael Clifford and Shuai Wang"
date: '2023-09-22'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)
library(mapview)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")
```

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("730e334cd384273fdcda9984e41c6444518d82d6", overwrite = TRUE)
```



```{r results='hide'}
#Get census tract data
tracts19 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E", "B25058_001E",
                        "B06012_002E"), 
          year=2019, state=25,
          county=025, geometry=TRUE) %>% 
  st_transform('ESRI:102686')

tracts09 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E", "B25058_001E",
                        "B06012_002E"), 
          year=2009, state=25,
          county=025, geometry=TRUE) %>% 
  st_transform('ESRI:102686')
```







### Working with ACS Data

```{r}
# Let's "spread" the data into wide form

tracts19 <- 
  tracts19 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         FemaleBachelors = B15001_050, 
         MaleBachelors = B15001_009,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         TotalPoverty = B06012_002)

tracts09 <- 
  tracts09 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         FemaleBachelors = B15001_050, 
         MaleBachelors = B15001_009,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         TotalPoverty = B06012_002)


# Let's create new rate variables using mutate

tracts19 <- 
  tracts19 %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2019") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)

tracts09 <- 
  tracts09 %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2009") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)
```


```{r}
# Combine the tracts from the two years
allTracts <- rbind(tracts09,tracts19)
```

### Wrangling Transit Open Data

```{r}

mbta <- st_read("https://hub.arcgis.com/datasets/massgis::mbta-rapid-transit-stops.geojson")

# Remove Silver Line Stops
mbta <- mbta %>% filter(LINE != "SILVER")

mbta <- mbta %>% st_transform(st_crs(tracts19))
```


```{r}
#plot the stops
ggplot() + 
  geom_sf(data=st_union(tracts19)) +
  geom_sf(data=mbta, 
          aes(colour = LINE), 
          show.legend = "point", size= 2) +
  
  labs(title="MBTA Stops", 
       subtitle="Boston, MA", 
       caption="Figure 2.5") +
  mapTheme()
```



```{r}
#create .5 mile buffer from stations
stopBuffer <- st_buffer(mbta, 2640)

stopUnion <- st_union(st_buffer(mbta, 2640))

mbtaBuffers <- 
  rbind(
     stopBuffer %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
     stopUnion %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))
```

Let's examine both buffers by making a "[small
multiples](https://www.juiceanalytics.com/writing/better-know-visualization-small-multiples)"
plot. This type of plot shows multiple views of the same data or
geographic area. We accomplish this by using the `facet_wrap()`
{ggplot2} function on the *Legend* column.

```{r}
ggplot() +
  geom_sf(data=mbtaBuffers) +
  geom_sf(data=mbta, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 2.6") +
  mapTheme()
```

Create an sf object with ONLY the unioned buffer from the `septaBuffers`
spatial data frame.

```{r}
buffer <- filter(mbtaBuffers, Legend=="Unioned Buffer")
```

### Spatial Intersection of tracts and buffers

#### Spatial intersection with with `st_centroid()` on polygon centroids

Finally, we are using centroids, or the point at the center of a polygon, to select whether a polygon from `tracts16` intersects the `buffer`. The steps here are:

1.  convert tract polygons to centroids and to the spatial intersect to buffers using the `selection1` approach from above. The resulting geometry type is now a point and not a polygon

2.  drop the geometry so that is becomes a regular non-spatial data frame

3.  join the original `tracts16` data frame using the `GEOID` as the join field. Note that only the `GEOID` column is selected, however the polygon spatial features are also joined.

4.  `st_sf()` is a function to turn it back to an {sf} spatial object

5.  finally the `TotalPop` column is selected and a new field of `Selection_Type` is added

Note: you will get a warning about "st_centroid assumes attributes... don't worry about it.

```{r}
selectCentroids <-
  st_centroid(tracts19)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts19, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")
```



```{r}
ggplot() +
  geom_sf(selectCentroids, mapping = aes(fill = TotalPop)) +
  geom_sf(data=mbta, show.legend = "point") +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()
```

### Indicator Maps


```{r}
#join TOD and non-TOD tracts by centroid method
allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2009", MedRent * 1.19, MedRent)) 

```

```{r}
#plot TOD vs non-TOD areas
ggplot() +
  geom_sf(data=allTracts.group, mapping = aes(fill = TOD)) +
  mapTheme()
```

2. Visualizations

```{r}
ggplot() +
  geom_sf(data = allTracts.group, aes(fill = q5(MedHHInc))) +
  facet_wrap(~year) +
  scale_fill_manual(values = palette5,
                    labels = qBr(allTracts.group, "MedHHInc"),
                    name = "Med. HH Income\n(Quintile Breaks)") +
  labs(title = "Med. HH Income", subtitle = "Boston; 2009-2019") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))
```

### TOD Indicator Plots (3. Bar Plots)

Let's create small multiple plots We use the "gather" command (look this
one up please) To go from wide to long Why do we do this?? Notice we can
"pipe" a ggplot call right into this operation!

```{r}
#Create summary table first
allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Rent = mean(MedRent, na.rm = T),
            Population = mean(TotalPop, na.rm = T),
            Percent_White = mean(pctWhite, na.rm = T),
            Percent_Bach = mean(pctBachelors, na.rm = T),
            Percent_Poverty = mean(pctPoverty, na.rm = T))
```

```{r}
allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Variable, scales = "free", ncol=5) +
  scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
  labs(title = "Indicator differences across time and space") +
  plotTheme() + theme(legend.position="bottom")
```

### TOD Indicator Tables

```{r}
#bad table
kable(allTracts.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.2")
```


(4.Table)

```{r}
allTracts.Summary %>%
  unite(year.TOD, year, TOD, sep = ": ", remove = T) %>%
  gather(Variable, Value, -year.TOD) %>%
  mutate(Value = round(Value, 2)) %>%
  spread(year.TOD, Value) %>%
  kable() %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.3")
```




### Example of using `multipleRingBuffer()` function

There is a more comprehensive tutorial on the usage of this function (and the bugs you might encounter) in the Github folder!

```{r}
allTracts <- rbind(tracts20,tracts16)
septa_MRB <- multipleRingBuffer(st_union(septaStops), 47520, 2640)

allTracts.rings <-
  st_join(st_centroid(dplyr::select(allTracts, GEOID, year)),
          septa_MRB) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(allTracts, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles



  ggplot() +
    geom_sf(data=septa_MRB) +
    geom_sf(data=septaStops, size=1) +
    geom_sf(data=st_union(tracts20), fill=NA, size=1.2) +
    labs(title="Half mile buffers") +
    mapTheme()

```

